---
title: "EPI590R Final Project"
author: "Ernest Koh"
format:
  html:
   theme: cayman
execute:
 echo: false
 cache: true
 warning: false
editor: visual
---

## **Factors Associated with Survival in the Titanic Disaster**

**Introduction:** On the 10th of April, 1912, the RMS Titanic set out on its maiden voyage across the Atlantic Ocean, carrying 2,223 passengers. On the 14th of April, it hit an iceberg and sank. In this analysis, we examine factors associated with the probability of survival in the Titanic disaster, including age, sex and ticket class using data available on 1309 of the passengers.

```{r}
#reading dataset
#| echo: false
library(readxl)
library(tidyverse)
library(gtsummary)
titanic_data <- read_xlsx(here::here("data", "titanic 2023.xlsx"))|>
	mutate(Survived_cat = factor(Survived, labels = c("(0) No", "(1) Yes")),
				 TicketClass_cat = factor(TicketClass, labels = c("First", "Second", "Third")))
```

**Descriptive Statistics**

```{r}
#| label: tbl-one
#| tbl-cap: "Descriptive Statistics"
table1 <- tbl_summary(
  titanic_data,
  by = Sex,
  include = c(Age, TicketClass, Survived),
  label = list(
  	Age ~ "Age in Years",
  	TicketClass ~ "Ticket Class",
  	Survived ~ "Survived"
  ),
  missing_text = "Missing") |>
  add_p(test = list(all_continuous() ~ "t.test", 
                    all_categorical() ~ "chisq.test")) |> 
  add_overall(col_label = "**Total**") |> 
  bold_labels() |> 
  modify_footnote(update = everything() ~ NA) |> 
  modify_header(label = "**Variable**", p.value = "**P**")
table1

survived <- list(number_survival = sum(titanic_data$Survived == 1, na.rm = TRUE))
died <- list(number_died = sum(titanic_data$Survived == 0, na.rm = TRUE))

survived_f <- inline_text(table1, variable = "Survived", column = "female")
survived_m <- inline_text(table1, variable = "Survived", column = "male")


```
@tbl-one shows the initial descriptive statistics of the Titanic dataset. The number of passengers who survived the titanic disaster is `r survived` and the number of passengers who died is `r died`. A greater proportion of female (`r survived_f`) than male (`r survived_m`) passengers survived the disaster.

**Univariate Logistic Regression Analysis**

```{r}
#| echo: false
#| label: tbl-two
#| tbl-cap: "Univariate Logistic Regression"
table2 <- tbl_uvregression(
  titanic_data, 
  y = Survived_cat,
  include = c(Age, TicketClass_cat, Sex),
  method = glm,
  method.args = list(family = binomial()),
  exponentiate = TRUE,
  label = list(Age ~ "Age",
		Sex ~ "Sex",
		TicketClass_cat ~ "Ticket Class"
		))
table2

uniOR_report <- inline_text(table2, variable = "TicketClass_cat", level = "Second")
```
@tbl-two shows the results from Univariate Logistic Regression.The odds of passengers with second class ticket surviving the Titanic disaster is `r uniOR_report` as compared to passengers with first class ticket.

**Multivariate Logistic Regression Analysis**

```{r}
#| echo: false
#| label: tbl-three
#| tbl-cap: "Multivariate Logistic Regression"
logistic_model <- glm(Survived_cat ~ Age + Sex + TicketClass_cat, 
                      data = titanic_data, family = binomial())
or_table <- tbl_regression(
	logistic_model,
	exponentiate = TRUE,
	label = list(
		Age ~ "Age",
		Sex ~ "Sex",
		TicketClass_cat ~ "Ticket Class"
))
or_table
or_report <- inline_text(or_table, variable="Sex", level = "male")
```
@tbl-three shows the results from Multivariate Logistic Regression. The odds of male passengers surviving the titanic disaster is `r or_report` as compared to female passengers controlling for Age and Ticket Class.

**OR Estimate Plots**

```{r}
#| echo: false
#| include: false
library(broom)
summary(logistic_model)
glance(logistic_model)
# this chunk is for personal review
```

```{r}
#| echo: false
#| label: fig-one
#| fig.cap: "Forest Plot of Estimate of the Odds Ratio"
options(digits = 3)
summary(logistic_model)$coefficients

tidy(logistic_model, conf.int = TRUE, exponentiate = TRUE) |>
	tidycat::tidy_categorical(logistic_model, exponentiate = TRUE) |>
	dplyr::slice(-1) |> 
	ggplot(mapping = aes(x = level, y = estimate,
											 ymin = conf.low, ymax = conf.high)) +
	geom_point() +
	geom_errorbar() +
  facet_grid(cols = vars(variable), scales = "free", space = "free") +
  scale_y_log10()

```
@fig-one shows the forest plot of the point estimate and 95% CI of the Odds Ratio.

```{r}
#| echo: false
#| include: false
#Getting the reciprocal odds ratio, a very simple function
x <- 0.89
1/x 
reciprocalOR <- function(x) {
  1/x}
reciprocalOR(0.46)

#Proportion function
y <- c(0,1,1,0)
multiplier <- 100
multiplier * sum(y)/length(y)

prop <- function(y, multiplier){
  n <- length(y)
  prop_val <- multiplier * sum(y)/n
  return(prop_val)
}
prop_f <- prop(y = titanic_data$Sex == "female", multiplier = 100)
options(digits = 3)

emo::ji("dog") #just for fun
```
**Conclusion**
Overall, the proportion of female passengers on board the Titanic is `r prop_f`. There is a statistically significant association between Survival Status and Sex, adjusted for Age and Ticket Class (p<0.001). The odds of surviving the titanic disaster for Male passengers are 0.08 (95% CI 0.06 - 0.11) times the odds for female passengers, adjusting for Age and Ticket Class. 
